Phylogenetic inference of contact network parameters with approximate Bayesian computation

### 2016-07-05 doc/defence/posterior1.tex
Presentation updates

\documentclass{beamer}
\usepackage{tikz}
\usepackage{gillius}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage{multirow}
\usepackage{qrcode}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}

% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}

\graphicspath{{../figures/}}
\newcommand{\tablepath}{../tables}

\title{Phylogenetic inference of contact network parameters with kernel approximate Bayesian computation}
\author[RMM, RHL \& AFYP]{Rosemary M McCloskey$^1$ \and Richard H Liang$^1$ \and Art FY Poon$^{1,2}$}
\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
\date{HIV Dynamics \& Evolution, Woods Hole, USA, April 25, 2016}

\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}

\addbibresource{papers.bib}

\begin{document}
\setbeamercolor{background canvas}{bg=white}
\renewcommand{\footnotesize}{\tiny}
\definecolor{red}{RGB}{228,26,28}
\definecolor{blue}{RGB}{55,126,184}
\definecolor{green}{RGB}{77,175,74}
\definecolor{purple}{RGB}{152,78,163}

\maketitle

\begin{frame}{Most epidemiological models assume homogeneous mixing}
  \begin{columns}
    \begin{column}{0.6\textwidth}
      \centering
      \includegraphics[scale=0.6]{compartments}
      \vspace{1cm}

      \uncover<3->{\includegraphics[width=\textwidth]{sir-trajectories}}
    \end{column}
    \begin{column}{0.4\textwidth}
      \vspace{-2cm}
      \begin{itemize}
        \setlength{\itemsep}{24pt}
        \uncover<2->{
        \item Often provide a reasonable approximation in practice.
        }
        \uncover<3->{
        \item Can be inaccurate when substantial contact heterogeneity exists.
        }
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Network models capture contact heterogeneity}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{contactnet-empty}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
        \setlength{\itemsep}{12pt}
          \pause
        \item May offer more accurate predictions for highly structured
          populations.
          \pause
        \item Network parameters may be of interest for their own sake,
          \textit{e.g.} are there superspreaders?
          \pause
        \item Extremely difficult to estimate in practice.
      \end{itemize}
    \end{column}
  \end{columns}

\end{frame}

\begin{frame}{Contact networks shape transmission trees}
  \includegraphics[width=\textwidth]{contactnet}

  \begin{itemize}
      \pause
    \item Transmission trees in turn shape viral phylogenies.
      \pause
    \item \textbf{Aim:} estimate contact network parameters from viral
      phylogenies.
  \end{itemize}
\end{frame}

%\begin{frame}{Preferential attachment generates scale-free networks}
%  \includegraphics[width=\textwidth]{pa-example}
%  \begin{itemize}
%    \item Sexual networks are often scale-free: degree distributions appear
%      follow a power law ($\Pr(\text{degree} = d) \propto d^{-\gamma}$).
%      \pause
%    \item Preferential attachment models generate scale-free networks.
%  \end{itemize}
%\end{frame}

\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{minipage}[p][\textheight][t]{\textwidth}
        \centering
        \only<1>{\includegraphics[page=1, width=\textwidth]{pa.pdf}}
        \only<2>{\includegraphics[page=2, width=\textwidth]{pa.pdf}}
        \only<3>{\includegraphics[page=3, width=\textwidth]{pa.pdf}}
        \only<4>{\includegraphics[page=4, width=\textwidth]{pa.pdf}}
        \only<5>{\includegraphics[page=5, width=\textwidth]{pa.pdf}}
        \only<6>{\includegraphics[page=6, width=\textwidth]{pa.pdf}}
      \end{minipage}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{minipage}[t][\textheight][t]{\textwidth}
      \begin{itemize}
        \uncover<1->{ \item Start with a small number of connected nodes. }
        \uncover<2->{ \item Attach new nodes with $m$ edges. }
        \uncover<3->{ \item Other endpoints of degree $d$ are chosen with probability $\propto d^{\alpha} + 1$. }
        \uncover<5->{ \item Continue until network has $N$ nodes. }
        \uncover<6->{ \item Also consider the prevalence $I$. }
      \end{itemize}
      \end{minipage}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Do network parameters measurably affect tree shape?}
  \only<1>{
    \includegraphics[height=1in, trim=0 2in 4in 0, clip]{kernel-idea}
  }
  \only<2>{
    \includegraphics[height=1in, trim=0 2in 2in 0, clip]{kernel-idea}
  }
  \only<3->{
    \includegraphics[height=1in, trim=0 2in 0 0, clip]{kernel-idea}
  }
  \begin{itemize}
    \setlength{\itemsep}{0pt}
    \item Generate networks under different parameter values (number of nodes
      $N$, number of edges per vertex $m$, pereferential attachment power
      $\alpha$).
    \uncover<2->{
    \item Simulate epidemic over each network until $I$ nodes are infected.
    }
    \uncover<3->{
    \item Randomly subsample to form transmission trees.
    }
    \uncover<4->{
    \item Compare trees pairwise using tree kernel\footfullcite{poon2013mapping}.
    }
  \end{itemize}
\end{frame}

\begin{frame}{Preferential attachment power $\alpha$ affects tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 2.7in 3.3in 0, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-alpha-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Prevalence $I$ affects tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 2.7in 0in 0, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-I-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Number of edges per vertex $m$ does not affect tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 0.3in 3.3in 2.2in, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-m-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Number nodes $N$ modestly affects tree shape}
  \begin{tikzpicture}[remember picture, overlay]
    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 0.3in 0 2.2in, clip]{kernel-kpca.pdf}};
    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-N-tree.pdf}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Simulated trees can be used to fit a network model}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
    \uncover<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
    \uncover<5->{\vspace{0.5cm}\centerline{\textbf{kernel approximate Bayesian computation (kernel-ABC)}}}
  \end{minipage}
\end{frame}

\begin{frame}{Example simulation shows $\alpha$ and $I$ can be reconstructed}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \begin{center}
    preferential attachment power \hfill prevalence
    \only<1>{
      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
    }
    \only<2>{
      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}

      \vspace{-0.25cm}
      edges per vertex \hfill total nodes
    }
    \end{center}
  \end{minipage}
\end{frame}

\begin{frame}{Maximum \textit{a posteriori} estimates are accurate for $\alpha$ and $I$}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \vspace{-0.5cm}
    preferential attachment power \hfill \uncover<2>{edges per vertex}

    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m2.pdf}}

    prevalence \hfill \uncover<2>{total nodes}
  \end{minipage}
\end{frame}

\begin{frame}{Real world HIV datasets exhibit network heterogeneity}
  \uncover<2->{
    \centerline{\includegraphics[height=0.8\textheight]{realdata-hpd-bc}}
  }

%  \scriptsize
%    \begin{tabular}{ccccc}
%      Reference & Sequences ($n$) & Location & Risk group & Gene \\
%      \hline
%      \color{magenta} \textcite{novitsky2013phylogenetic} &
%        \color{magenta} \multirow{2}{*}{180} &
%        \color{magenta} \multirow{2}{*}{Mochudi, Botswana} & 
%        \color{magenta} \multirow{2}{*}{HET} &
%        \color{magenta} \multirow{2}{*}{\textit{env}} \\ 
%      \color{magenta} \textcite{novitsky2014impact} \\
%      \color{blue} \textcite{cuevas2009hiv} & 
%        \color{blue} 287 & 
%        \color{blue} Basque Country, Spain & 
%        \color{blue} mixed & 
%        \color{blue} \textit{pol} \\
%      \color{cyan} \textcite{li2015hiv} & 
%        \color{cyan} 280 & 
%        \color{cyan} Shanghai, China & 
%        \color{cyan} MSM & 
%        \color{cyan} \textit{pol} \\
%      \color{green} unpublished & 
%        \color{green} 399 & 
%        \color{green} British Columbia, Canada & 
%        \color{green} IDU & 
%        \color{green} \textit{pol} \\
%      \color{orange} \textcite{wang2015targeting} & 
%        \color{orange} 173 & 
%        \color{orange} Beijing, China & 
%        \color{orange} MSM & 
%        \color{orange} \textit{pol} \\
%      \color{red} \textcite{niculescu2015recent} & 
%        \color{red} 136 & 
%        \color{red} Romaina & 
%        \color{red} IDU & 
%        \color{red} \textit{pol} \\
%      \hline
%    \end{tabular}
%  \normalsize
\end{frame}

\begin{frame}{Conclusions}
  \begin{itemize}
    \item Kernel-ABC is the first phylodynamic method to fit contact network
      models to phylogenetic data.
      \pause
    \item The preferential attachment power of the Barab\'asi-Albert network
      model, which is challenging to estimate by traditional epidemiological
      methods, can be estimated with kernel-ABC.
      \pause
    \item The networks underlying real epidemics are heterogeneous,
      underscoring the importance of considering network structure in
      phylodynamic analyses.
      \pause
  \end{itemize}
  \centerline{\qrcode{github.com/rmcclosk/netabc} github.com/rmcclosk/netabc}
\end{frame}

\begin{frame}{Acknowledgements}
  \begin{columns}
    \begin{column}{0.6\textwidth}

      \textbf{BC Centre for Excellence in HIV/AIDS}

      Art Poon

      Jeff Joy

      Richard Liang

      Thuy Nguyen

      P. Richard Harrigan

      \hfill\\
      \textbf{University of British Columbia}

      Sarah Otto

      Alexandre Bouchard-C\^ot\'e

      \vfill
      \vspace{0.5cm}
      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
      $\qquad$
      \includegraphics[width=2cm]{logos/genomequebec}
    \end{column}
    \begin{column}{0.4\textwidth}
      \centering

      \includegraphics[width=3cm]{logos/cfe}
      \vspace{0.5cm}

      \includegraphics[width=3cm]{logos/cihr}

      \includegraphics[width=3cm]{logos/genomebc}

      \includegraphics[width=3cm]{logos/bmgf}
      \vspace{0.5cm}

      \includegraphics[width=3cm]{logos/btp}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Classifiers for network parameters}
  \vspace{-1cm}
  \begin{itemize}
    \item kSVR using tree kernel
    \item SVR using normalized
      lineages-through-time\footfullcite{janzen2015approximate}
    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
  \end{itemize}

  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
\end{frame}

\begin{frame}{Sequential Monte Carlo for ABC}
  \begin{tikzpicture}
    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
  \end{tikzpicture}
\end{frame}

\end{document}

>>>>

\begin{frame}{Marginal posterior distributions for $\alpha$ show least spread}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \begin{center}
    preferential attachment power \hfill prevalence
      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}

      \vspace{-0.25cm}
      edges per vertex \hfill total nodes
    \end{center}
  \end{minipage}
\end{frame}

### 2016-07-05 doc/defence/presentation.tex
Updates to presentation
Presentation updates

\documentclass{beamer}
\usepackage{tikz}
\usepackage{gillius}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage{multirow}
\usepackage{qrcode}
\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
\usepackage{hyperref}

@@

\maketitle
\input{contactnet}
\input{netmodel}
\input{why}
\input{ba}
\input{phylonet}
\input{transtree}
\input{virusphylo}
\input{inference}
\input{identifiable}

@@

\begin{frame}{yes they do}
  \centerline{\includegraphics[height=0.8\textheight]{kernel-kpca}}
\end{frame}

\begin{frame}{Simulated trees can be used to fit network models}
  \centerline{\includegraphics[width=0.9\textwidth]{abc-smc}}

  \centerline{animate this?}
\end{frame}

\begin{frame}{Posterior mean estimates are more accurate for $\alpha$ and $I$}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \vspace{-0.5cm}
    preferential attachment power \hfill \uncover<2>{edges per vertex}

    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m3.pdf}}
    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m3.pdf}}

    prevalence \hfill \uncover<2>{total nodes}
  \end{minipage}
\end{frame}

\begin{frame}{Example simulation\ldots}
  1d posterior goes here
  \pause
%  \begin{minipage}[p][\textheight][t]{\textwidth}
%    \begin{center}
%    preferential attachment power \hfill prevalence
%    \only<1>{
%      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
%    }
%    \only<2>{
%      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
%
%      \vspace{-0.25cm}
%      edges per vertex \hfill total nodes
%    }
%    \end{center}
%  \end{minipage}
\end{frame}

\begin{frame}{Example simulation continued, $N$ and $I$ confounded}
  2d posterior goes here
  %\includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}
\end{frame}

\begin{frame}{Application to real world HIV datasets}
  \centerline{\input{\tablepath/realdata-bc-short}}
\end{frame}

>>>>

\documentclass{beamer}
\usepackage{tikz}
\usepackage{gillius}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage{multirow}
\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
\usepackage{hyperref}

@@

\maketitle

%\begin{frame}{Outline}
%    \tableofcontents
%\end{frame}

%\section{Background and Motivation}
%\input{contactnet}
%\input{phylo}
%\input{objects}
%\input{transtree}
%\input{virusphylo}
%\input{inference}
%%\input{netmodel}
%\input{why}

%\section{The Barab\'asi-Albert (BA) Contact Network Model}
%\input{ba}
%\input{identifiable}
%\input{kpca}

%\section{\textit{Netabc}: Phylogenetic Reconstruction of Contact Network Parameters with Approximate Bayesian Computation}
%\input{smc}
%\input{posterior1}
%\input{posterior2}
%\input{pointest}
%\input{pointest2}

\section{Application to real HIV epidemics}
\input{data}

\section{Conclusions}

@@

### 2016-07-05 doc/defence/test.tex
Presentation updates

\maketitle

%\begin{frame}{Outline}
%    \tableofcontents
%\end{frame}

%\section{Background and Motivation}
%\input{contactnet}
%\input{phylo}
%\input{objects}
%\input{transtree}
%\input{virusphylo}
%\input{inference}
%%\input{netmodel}
%\input{why}

%\section{The Barab\'asi-Albert (BA) Contact Network Model}
%\input{ba}
%\input{identifiable}
%\input{kpca}

%\section{\textit{Netabc}: Phylogenetic Reconstruction of Contact Network Parameters with Approximate Bayesian Computation}
%\input{smc}
%\input{posterior1}
%\input{posterior2}
%\input{pointest}
%\input{pointest2}

\section{Application to real HIV epidemics}
\input{data}

\section{Conclusions}

@@

\begin{frame}{HIV networks have sub-linear preferential attachment}
    Also, higher PA for IDU networks.

    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.5\textwidth]{realdata-hpd-bc}
    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.5\textwidth]{realdata-hpd-bc}
\end{frame}

\begin{frame}{Conclusions}
  \begin{itemize}
    \item We developed a phylodynamic method to fit contact network models to
      phylogenetic data.
      \pause
    \item The preferential attachment power of the Barab\'asi-Albert network
      model, which is challenging to estimate by traditional epidemiological
      methods, can be estimated with ABC.
      \pause
    \item The networks underlying real epidemics are heterogeneous,
      underscoring the importance of considering network structure in
      phylodynamic analyses.
  \end{itemize}
\end{frame}

\begin{frame}{Acknowledgements}
  \begin{columns}
    \begin{column}{0.6\textwidth}

      \textbf{BC Centre for Excellence in HIV/AIDS}

      Art Poon

      Jeff Joy

      Richard Liang

      Thuy Nguyen

      P. Richard Harrigan

      \hfill\\
      \textbf{University of British Columbia}

      Sarah Otto

      Alexandre Bouchard-C\^ot\'e

      \vfill
      \vspace{0.5cm}
      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
      $\qquad$
      \includegraphics[width=2cm]{logos/genomequebec}
    \end{column}
    \begin{column}{0.4\textwidth}
      \centering

      \includegraphics[width=3cm]{logos/cfe}
      \vspace{0.5cm}

      \includegraphics[width=3cm]{logos/cihr}

      \includegraphics[width=3cm]{logos/genomebc}

      \includegraphics[width=3cm]{logos/bmgf}
      \vspace{0.5cm}

      \includegraphics[width=3cm]{logos/btp}
    \end{column}
  \end{columns}
\end{frame}

>>>>

\input{objects2}

@@
